---
title: "fodreg: DAG networks of Gulf of Bothnia basins (annual data)"
subtitle: "Dynamic Bayesian networks with dbnR – drivers of primary and bacterial production"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    embed-resources: true
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
params:
  data_file: "AnnualBasinDataGOB94t23.xlsx"
  size: 2          # number of time slices: 2 = t_0 and t_1 (one step back = 1 year here)
  intra: true      # allow arcs within the same time slice (same-year effects)
  detrend: false   # remove linear trend (by year) per variable and basin before learning
  n_boot: 200      # bootstrap replicates for arc-strength section
  threshold: 0.5   # arc-strength threshold for the averaged network
---

# Aim

Infer which environmental variables (and zooplankton) lead to the two response variables

- **Phytoplankton production** (`PhytProd`)
- **Bacterioplankton production** (`BaGroBm`, biomass production; swap to `BaGroCel` for cell growth if preferred)

using a dynamic Bayesian network (DBN) in the R package **dbnR**, with one network per basin:
Bothnian Bay, Bothnian Sea and the Öre estuary. This first version uses the **annual** values
(1994–2023). The code is written so the same functions can be re-used on seasonal data later (see the last section).

# Setup

```{r setup}
library(readxl)
library(data.table)
library(dplyr)
library(ggplot2)
library(bnlearn)
library(dbnR)
library(visNetwork)

set.seed(42)
```

## Choices

```{r choices}
# Basin codes in the file are A, B and C. The metadata sheet does not say which is which.
# Mapping below is an ASSUMPTION based on the salinity gradient (A ~3, B ~4, C ~5 PSU)
# -> CHECK WITH THE MAIN AUTHOR.
basin_map <- c(A = "Bothnian Bay", B = "Bothnian Sea", C = "Oere estuary")

# Responses
responses <- c(PhytProd = "Phytoplankton production",
               BaGroBm  = "Bacterial biomass production")

# Prioritised predictors (PLS selection by co-author)
predictors <- c(TotP     = "Total phosphorus",
                PO4      = "Phosphate",
                Cladocer = "Cladocerans",
                Copepoda = "Copepoda",
                Salinity = "Salinity",
                SiO4     = "Silicate",
                GlobIrra = "Global irradiation",
                DOC      = "DOC",
                TotN     = "Total nitrogen")

vars <- c(names(responses), names(predictors))
labels <- c(responses, predictors)
```

# Data

```{r data}
raw <- read_excel(params$data_file, sheet = 1,
                  na = c("", "NA", "#NULL!")) |>
  as.data.table()

dat <- raw[, c("Seabasin", "Year", vars), with = FALSE]
dat[, Basin := basin_map[Seabasin]]
setorder(dat, Basin, Year)

# Missing values per variable and basin
miss <- dat[, lapply(.SD, function(x) sum(is.na(x))), by = Basin, .SDcols = vars]
knitr::kable(miss, caption = "Number of missing years per variable and basin")
```

Years with missing values (these only remove the time-step pairs that involve them, see below):

```{r missing-years}
dat[, .(years_with_NA = paste(Year[!complete.cases(.SD)], collapse = ", ")),
    by = Basin, .SDcols = vars] |>
  knitr::kable()
```

## Overview of the time series

```{r ts-plot}
#| fig-height: 9
long <- melt(dat, id.vars = c("Basin", "Year"), measure.vars = vars,
             variable.name = "var", value.name = "value")
long[, var_label := factor(labels[as.character(var)], levels = labels)]

ggplot(long, aes(Year, value, colour = Basin)) +
  geom_line(na.rm = TRUE) +
  facet_wrap(~var_label, scales = "free_y", ncol = 3) +
  theme_bw() + theme(legend.position = "bottom")
```

# Preprocessing

For each basin separately:

1. optionally remove a linear trend (`params$detrend`),
2. standardise every variable to mean 0 and SD 1 (z-score, within basin),
3. fold the series into time slices (`t_0` = current year, `t_1` = previous year) with `dbnR::fold_dt()`,
4. drop rows of the folded table that contain `NA`.

Folding is done on the *complete* year sequence (with `NA`s kept in place) so that lags are never
formed across a gap; only the affected pairs are dropped afterwards.

```{r preprocess}
prep_basin <- function(d, vars, size = 2, detrend = FALSE) {
  d <- copy(d)
  setorder(d, Year)
  stopifnot(all(diff(d$Year) == 1))   # consecutive time steps required for lags

  x <- as.data.table(d[, ..vars])

  if (detrend) {
    x <- x[, lapply(.SD, function(v) as.numeric(resid(lm(v ~ d$Year, na.action = na.exclude))))]
  }

  # z-score within basin (NA-safe)
  x <- x[, lapply(.SD, function(v) as.numeric(scale(v)))]

  f <- dbnR::fold_dt(x, size = size)
  f <- na.omit(f)
  list(z = x, folded = f)
}

basins <- unname(basin_map)
prep <- lapply(setNames(basins, basins), function(b) {
  prep_basin(dat[Basin == b], vars, size = params$size, detrend = params$detrend)
})

data.frame(Basin = basins,
           n_years = sapply(basins, function(b) nrow(dat[Basin == b])),
           n_rows_used = sapply(prep, function(p) nrow(p$folded)),
           n_nodes = sapply(prep, function(p) ncol(p$folded)),
           row.names = NULL) |>
  knitr::kable(caption = "Rows available for structure learning after folding")
```

::: {.callout-warning}
## Small sample size
With roughly 24–29 usable rows and 22 nodes (11 variables × 2 time slices), the learned structure will be
unstable and sparse. Treat single arcs with caution and use the bootstrap section below to see which arcs are
robust. Annual series with trends (salinity, DOC/browning, irradiation) can also create apparent links
between variables that simply share a trend. Compare `detrend: true` against `false`.
:::

# Structure learning with dbnR

`learn_dbn_struc()` uses the dynamic max–min hill-climbing algorithm (`dmmhc`) by default. Arcs can only go
from `t_1` to `t_0`; with `intra: true` arcs within `t_0` (same-year effects) are allowed as well.
With annual data, same-year effects are probably the most interesting ones, whereas the lag-1 arcs often
mainly capture autocorrelation.

```{r learn}
learn_net <- function(folded, size, intra) {
  dbnR::learn_dbn_struc(dt = NULL, size = size, method = "dmmhc",
                        f_dt = folded, intra = intra)
}

nets <- lapply(prep, function(p) learn_net(p$folded, params$size, params$intra))
```

```{r helpers}
# Parents of a node (arcs pointing into it)
arcs_into <- function(net, targets) {
  a <- as.data.frame(bnlearn::arcs(net))
  if (nrow(a) == 0) return(a)
  a[a$to %in% targets, , drop = FALSE]
}

target_nodes <- paste0(names(responses), "_t_0")

plot_net <- function(net, title) {
  nodes <- data.frame(id = bnlearn::nodes(net), stringsAsFactors = FALSE)
  base <- sub("_t_[0-9]+$", "", nodes$id)
  slice <- sub("^.*_t_", "t_", nodes$id)
  nodes$label <- paste0(base, " (", slice, ")")
  nodes$group <- ifelse(nodes$id %in% target_nodes, "Response",
                        ifelse(slice == "t_0", "Predictor t_0", "Predictor t_1"))
  a <- as.data.frame(bnlearn::arcs(net))
  edges <- if (nrow(a)) data.frame(from = a$from, to = a$to, arrows = "to") else NULL

  visNetwork(nodes, edges, main = title, height = "600px", width = "100%") |>
    visGroups(groupname = "Response",      color = "#d95f02", shape = "box") |>
    visGroups(groupname = "Predictor t_0", color = "#1b9e77") |>
    visGroups(groupname = "Predictor t_1", color = "#7570b3") |>
    visLegend() |>
    visOptions(highlightNearest = list(enabled = TRUE, degree = 1)) |>
    visPhysics(stabilization = TRUE)
}
```

## Bothnian Bay

```{r net-bay}
plot_net(nets[["Bothnian Bay"]], "Bothnian Bay")
```

```{r arcs-bay}
knitr::kable(arcs_into(nets[["Bothnian Bay"]], target_nodes), row.names = FALSE,
             caption = "Arcs into the response nodes, Bothnian Bay")
```

## Bothnian Sea

```{r net-sea}
plot_net(nets[["Bothnian Sea"]], "Bothnian Sea")
```

```{r arcs-sea}
knitr::kable(arcs_into(nets[["Bothnian Sea"]], target_nodes), row.names = FALSE,
             caption = "Arcs into the response nodes, Bothnian Sea")
```

## Öre estuary

```{r net-ore}
plot_net(nets[["Oere estuary"]], "Öre estuary")
```

```{r arcs-ore}
knitr::kable(arcs_into(nets[["Oere estuary"]], target_nodes), row.names = FALSE,
             caption = "Arcs into the response nodes, Öre estuary")
```

## Fitted parameters (linear Gaussian)

Coefficients are in standardised units, so they are comparable between predictors within a basin.

```{r fit}
fits <- Map(function(net, p) {
  tryCatch(dbnR::fit_dbn_params(net, p$folded, method = "mle-g"),
           error = function(e) NULL)
}, nets, prep)

for (b in basins) {
  cat("\n###", b, "\n")
  fit <- fits[[b]]
  if (is.null(fit)) { cat("Parameter fitting failed.\n"); next }
  for (tn in target_nodes) {
    cf <- fit[[tn]]$coefficients
    cat("\n**", tn, "**\n\n")
    print(round(cf, 3))
  }
}
```

# Robustness: bootstrapped arc strength

A single hill-climbing run on ~25 observations is fragile. Here the same time-slice restrictions are imposed
as a blacklist in `bnlearn`, the structure is learned on bootstrap resamples, and the arcs are summarised by
how often they appear (`strength`) and how consistent their direction is (`direction`). This is not
`dbnR` itself but works on the same folded data.

```{r boot}
make_blacklist <- function(nodes, intra) {
  slice <- as.integer(sub("^.*_t_", "", nodes))
  bl <- expand.grid(from = nodes, to = nodes, stringsAsFactors = FALSE)
  s_from <- as.integer(sub("^.*_t_", "", bl$from))
  s_to   <- as.integer(sub("^.*_t_", "", bl$to))
  # nothing can be caused by the present or later than the effect: only older -> newer
  forbid <- s_from < s_to                       # t_0 -> t_1 (future causing past)
  forbid <- forbid | (s_from == s_to & s_to > 0) # no arcs within the past slice(s)
  if (!intra) forbid <- forbid | (s_from == 0 & s_to == 0)
  bl <- bl[forbid | bl$from == bl$to, ]
  bl
}

boot_one <- function(folded, intra, R, threshold) {
  df <- as.data.frame(folded)
  bl <- make_blacklist(names(df), intra)
  st <- boot.strength(df, R = R, algorithm = "hc",
                      algorithm.args = list(score = "bic-g", blacklist = bl))
  list(strength = st,
       avg = averaged.network(st, threshold = threshold))
}

boots <- lapply(prep, function(p) boot_one(p$folded, params$intra, params$n_boot, params$threshold))
```

```{r boot-tables}
for (b in basins) {
  st <- boots[[b]]$strength
  st <- st[st$to %in% target_nodes & st$strength >= 0.3 & st$direction >= 0.5, ]
  st <- st[order(st$to, -st$strength), ]
  cat("\n###", b, "\n\n")
  print(knitr::kable(st, row.names = FALSE, digits = 2,
                     caption = paste("Arcs into response nodes with strength >= 0.3 -", b)))
}
```

```{r boot-plots}
#| results: asis
for (b in basins) {
  cat("\n### Averaged network, ", b, " (threshold ", params$threshold, ")\n\n", sep = "")
  print(htmltools::tagList(plot_net(boots[[b]]$avg, paste(b, "- bootstrap-averaged"))))
}
```

# Extending to seasonal data

The functions above only assume a table with one row per consecutive time step. For seasonal data:

- Replace the `Year` column by a running time index (e.g. year + season; all seasons present, with `NA` in gaps) and
  make sure `diff()` of the index is constant, so that `t_1` always means "one season back".
- Standardise **within season** (or remove the seasonal cycle) *and* within basin, otherwise the seasonal cycle itself will
  dominate the correlations. Alternatively keep the seasonal cycle and add a season term (e.g. `sin`/`cos` of day of year) as an
  exogenous node that is only allowed to point to other nodes (use a whitelist/blacklist).
- Light, temperature and the productivity variables lag each other by weeks to months; consider `size = 3`
  or the number of seasons per year as a way of covering one year back.
- Keep in mind that the number of rows grows by roughly the number of seasons per year, which helps the
  stability problem considerably.

# Session info

```{r session}
sessionInfo()
```
